Multiway Spherical Clustering via Degree-Corrected\\ Tensor Block Models\footnote{Part of this paper was presented at 25th International Conference on Artificial Intelligence and Statistics (AISTATS).}

### 2022-01-20 tensor_clustering/tensor_clustering_arxiv/tensor_clustering_arxiv.tex
package

\documentclass[lettersize,onecolumn,journal]{IEEEtran}
\usepackage{amsmath,amsfonts}
\usepackage{algorithm}
\usepackage{array}
\usepackage[caption=false,font=normalsize,labelfont=sf,textfont=sf]{subfig}
\usepackage{textcomp}
\usepackage{stfloats}
\usepackage{url}
\usepackage{verbatim}
\usepackage{graphicx}
\usepackage{cite}
\hyphenation{op-tical net-works semi-conduc-tor IEEE-Xplore}
% updated with editorial comments 8/9/2021

@@

\usepackage{amsmath,amssymb,amsthm,bm,mathtools}
\usepackage{algorithm}
\usepackage{dsfont,multirow,hyperref,setspace,enumerate}
\hypersetup{colorlinks,linkcolor=black,citecolor=black,urlcolor={black}}
\usepackage{lscape}
\usepackage{afterpage}
\usepackage{hyperref}

@@

@@

\usepackage{fullpage} 
\usepackage{microtype}
\usepackage{wrapfig}
\mathtoolsset{showonlyrefs}

@@

\newcommand{\Mat}{\text{Mat}}

@@

\newcommand\Algphase[1]{%
\vspace*{-.7\baselineskip}\Statex\hspace*{\dimexpr-\algorithmicindent-2pt\relax}\rule{\columnwidth}{0.4pt}%
\Statex\hspace*{-\algorithmicindent}\textbf{#1}%
\vspace*{-.7\baselineskip}\Statex\hspace*{\dimexpr-\algorithmicindent-2pt\relax}\rule{\columnwidth}{0.4pt}%
}
\usepackage{caption}
\def\fixme#1#2{\textbf{\color{red}[FIXME (#1): #2]}}

@@

\title{Multiway Spherical Clustering via \\ Degree-Corrected Tensor Block Models}

\author{ Jiaxin Hu and Miaoyan Wang\\
University of Wisconsin - Madison
%IEEE Publication Technology,~\IEEEmembership{Staff,~IEEE,}
        % <-this % stops a space
%\thanks{This paper was produced by the IEEE Publication Technology Group. They are in Piscataway, NJ.}% <-this % stops a space
%\thanks{Manuscript received April 19, 2021; revised August 16, 2021.}
}

% The paper headers
%\markboth{Journal of \LaTeX\ Class Files,~Vol.~14, No.~8, August~2021}%
%{Shell \MakeLowercase{\textit{et al.}}: A Sample Article Using IEEEtran.cls for IEEE Journals}

%\IEEEpubid{0000--0000/00\$00.00~\copyright~2021 IEEE}
% Remember, if you use this you must call \IEEEpubidadjcol in the second
% column for its text to clear the IEEEpubid mark.

\maketitle

@@

\begin{IEEEkeywords}
tensor clustering, degree correction, statistical-computational efficiency, human brain connectome networks
\end{IEEEkeywords}

\section{Introduction}\label{sec:intro}
\IEEEPARstart{M}{ultiway} arrays have been widely collected in various fields including social networks \citep{anandkumar2014tensor}, neuroscience \citep{wang2017bayesian}, and computer science \citep{koniusz2016sparse}. Tensors effectively represent the multiway data and serve as the foundation in higher-order data analysis. One data example is from multi-tissue multi-individual gene expression study~\citep{wang2019three,hore2016tensor}, where the data tensor consists of expression measurements indexed by (gene, individual, tissue) triplets. Another example is \emph{hypergraph} network~\citep{ghoshdastidar2017uniform,ghoshdastidar2017consistency,ahn2019community,ke2019community} in social science. A $K$-uniform hypergraph can be naturally represented as an order-$K$ tensor, where each entry indicates the presence of $K$-way hyperedge among nodes (a.k.a.\ entities). In both examples, identifying the similarity among tensor entities is important for scientific discovery.

@@

Classical tensor block model suffers from drawbacks to model real world data in spite of the popularity. The key underlying assumption of block model is that all nodes in the same community are exchangeable; i.e., the nodes have no individual effects apart from the block effects. However, the exchangeability assumption is often non-realistic. Each node may contribute to the data variation by its own multiplicative effect. Such degree heterogeneity appears commonly in social networks. Ignoring the degree heterogeneity may seriously mislead the clustering results. For example, regular block model fails to model the member affiliation in the Karate Club network~\citep{bickel2009nonparametric} without addressing degree heterogeneity.

@@

\begin{itemize}
    \item \textit{Block model for clustering.} Block models such as stochastic block model (SBM) and degree-corrected SBM have been widely used for matrix clustering problems. The theoretical properties and algorithm performance for matrix block models have been well-studied~\citep{gao2018community}; see the review paper~\citep{abbe2017community} and the references therein. However, The tensor counterparts are relatively less understood.

    \item  \textit{Tensor block model.} The tensor block model (TBM, \cite{wang2019multiway, han2020exact, ghoshdastidar2017consistency}) is a higher-order extension of SBM, but it fails to allow degree heterogeneity. The Cartesian coordinates based analysis in \cite{han2020exact} is non-applicable to handle the extra complexity brought by degree heterogeneity. In contrast, our model addresses the degree heterogeneity, and we develop polar-coordinate based tools for the theoretical analysis.

    \item \textit{Degree-corrected block model.} The hypergraph degree-corrected block model (hDCBM) proposed by \cite{ke2019community, yuan2018testing} accounts for degree heterogeneity. However, the hDCBM is designed only for binary observations, and the proposed spectral algorithm in \cite{ke2019community} achieves sub-optimal polynomial clustering rate in higher-order scenarios. In constrast, our model allows discrete and continuous entries, and achieves \emph{exponentially} fast rate in clustering tasks. More importantly, to our best knowledge, we are the first to provide the statistical and computational limits analyses for the degree-corrected block model in tensor clustering. See Fig~\ref{fig:phase_axis} for overview of our results.

@@

Assume there exist $r$ disjoint communities among $p$ nodes, and the connection probabilities depend on the community assignments and node effects. 
Then, the equation~\eqref{eq:model_tensor} models $\bbE \tY$ with unknown degree heterogeneity $\mtheta$ and sub-Gaussianity parameter $\sigma^2 = 1/4$.

@@

\Algphase{Sub-algorithm 2: Angle-based iteration}
\INPUT Observation $\tY \in \bbR^{p \times \cdots \times p}$, initialization $z^{(0)} \colon [p]\to[r]$ from Sub-algorithm 1, iteration number $T$.
\For {$t = 0$ to $T-1$}
\State Update the block tensor $\tS^{(t)}$ via
$\tS^{(t)} (i_1,...,i_K)= \text{Ave} \{\tY(i_1,\ldots,i_K): z^{(t)}(i_k) = j_k, k \in [K]\}.$
\State Calculate reduced tensor $\tY^{\text{d}} \in \bbR^{p \times r \times \cdots \times r}$ via
\begin{align}
    &\tY^{\text{d}}(i,a_2,\ldots,a_K) 
    = 
\text{Ave}\{\tY(i,i_2,\ldots,i_K): z^{(t)}(i_k) = a_k, k \neq 1 \}.
\end{align}

@@

\begin{defn}[Locally linear stability] \label{def:stable}
Define the $\varepsilon$-neighborhood of $z$ by $\tN(z,\epsilon)=\{\bar z\colon \ell(\bar z, z)\leq \epsilon\}$. Let $\bar z\colon[p]\to [r]$ be a clustering function. We define two vectors associated with $\bar z$,
\begin{align}
    &\mp(\bar z)=(|\bar z^{-1}(1)|, \ldots,|\bar z^{-1}(r)|)^T, \quad \mp_{\mtheta}(\bar z)=(\onormSize{}{\mtheta_{\bar z^{-1}(1)}}_1,\ldots,\onormSize{}{\mtheta_{\bar z^{-1}(r)}}_1)^T.
\end{align}

@@

\begin{thm}[Error for angle-based iteration]\label{thm:refinement} Consider the setup as in Theorem~\ref{thm:initial}. Assume the local linear stability of degree holds in the neighborhood $\tN(z,\epsilon)$ for $\epsilon = \log^{-1}p$. Suppose 
$r=\tO(1)$ and $\text{SNR} \geq \tilde C p^{-K/2}\log p$ for some large constant $\tilde C$. Let $z^{(t)}$ denote the $t$-th iteration output in Sub-algorithm~\hyperref[alg:main]{2} with initialization $z^{(0)}$ from Sub-algorithm~\hyperref[alg:main]{1}. With probability going to 1, there exists a contraction parameter $\rho \in (0,1)$ such that 
\begin{align}\label{eq:final}
    \ell(z, \hat z^{(t+1)}) \lesssim &\ \KeepStyleUnderBrace{
   \text{SNR}^{-1}
    \exp\of{- \frac{p^{K-1}\text{SNR}}{r^{K-1}}}}_{\substack{\text{statistical error}} }+ \KeepStyleUnderBrace{ \rho^t \ell(z, z^{(0)}). }_{\substack{\text{computational error}}}
\end{align}
\end{thm}
From the conclusion~\eqref{eq:final}, we find that the iteration error is decomposed into two parts: statistical error and computational error. The statistical error is unavoidable with noisy data regardless $t$, whereas the computational error decays in an exponential rate as the number of iterations $t \rightarrow \infty$.

@@

{\bf Extension for general dTBMs.} Our two-stage algorithm is able to be extended for the general (asymmetric) dTBMs. Specifically, in the Sub-algorithm~\hyperref[alg:main]{1}, we make the following changes: (1)
Replace the matrcization $\mat(\tY)$ by $\mat_k(\tY)$; (2) Repeat the Steps 1-5 with mode-specified cluster number $r_k$; (3) Obtain the collection initialization $ \{z^{(0)}_k\}_{k=1}^K$. In the Sub-algorithm~\hyperref[alg:main]{2}, we make the following changes: (1) Take the collection $\{z^{(0)}_k\}_{k=1}^K$ as input, and update the block tensor $\tS^{(t)}$ with the collection $\{ z_k^{(t)}\}_{k=1}^K$, $
        \tS^{(t)}(i_1,\ldots, i_K) =  \text{Ave} \{\tY(i_1,\ldots,i_K): z^{(t)}_k(i_k) = j_k, k \in [K]\}$;
(2) Calculate reduced tensor $\tY^{\text{d}}_k$ for each mode via
\begin{align}
    &\tY^{\text{d}}_k(a_1,\ldots,a_{k-1}, i ,a_{k+1},\ldots,a_K) 
    = \text{Ave}\{\tY(i_1,\ldots,i_{k-1},i,i_{k+1},\ldots,i_K): z^{(t)}(i_j) = a_j, j \neq k \}; 
\end{align}

@@

\begin{table}[hbt]
\centering
    \begin{tabularx}{\textwidth}{c *{8}{Y}}
    \toprule
    Settings & \multicolumn{2}{c}{$p = 50, \sigma^2 = 0.25$} & \multicolumn{2}{c}{$p = 50, \sigma^2 = 1$} & \multicolumn{2}{c}{$p = 80, \sigma^2 = 0.25$} & \multicolumn{2}{c}{$p = 80, \sigma^2 = 1$}\\
    \cmidrule(lr){2-3} \cmidrule(l){4-5} \cmidrule(l){6-7} \cmidrule(l){8-9}
         True cluster number $r$ & 2 & 4 & 2  & 4 & 2  & 4 & 2 & 4  \\
         \midrule
         Estimated cluster number $\hat r$ &  2(0)  & 3.9(0.25)& 2(0)    & 3.1(0.52) & 2(0)    & 4(0)   & 2(0)    & 3.9(0.31)   \\
     \bottomrule
    \end{tabularx}
    \caption{Estimated cluster number given by BIC criterion under the low noise level $(\sigma^2 = 0.25)$ and high noise level $(\sigma^2 = 0.5)$ settings. Numbers in parentheses are standard deviations of $\hat r$ over 30 replications.}
    \label{tab:select}
\end{table}

@@

\newpage
\appendix

@@

\section*{Notation}
Before the proofs, we first introduce the notation used throughout the appendix and the generalized dTBM without symmetric assumptions. The parameter space and minimal gap assumption are also extended for the generalized dTBM.

@@

Consider an arbitrary $j \in [p_1]$ such that $z_1(j) = a$. Then for all the nodes $i \in  z_1^{-1}(a)$ in the same cluster of $j$, we have 
\begin{equation}
    \frac{\mX_{1,z_1(i):}}{\mX_{1,z_1(j):}} = \frac{\mX'_{1,z_1(i):}}{\mX'_{1,z_1(j):}}, \text{ which implies }  \frac{\theta_1(j)}{\theta_1(i)} = \frac{\theta'_1(j)}{\theta'_1(i)}.\label{eq:theta_uniq}
\end{equation}
Let $\theta'_1(j) = c\theta_1(j)$ for some positive constant $c$. By equation~\eqref{eq:theta_uniq}, we have $\theta'_1(i) = c \theta_1(i)$ for all $ i \in  z_1^{-1}(a)$. By the constraint $(\{z_k\}, \tS', \{\mtheta'_k\}) \in \tP(\{r_k\})$, we have 
\begin{equation}
    \sum_{j \in z^{-1}(a)} \theta'_1(j) = c \sum_{j \in z^{-1}(a)} \theta_1(j) = 1,
\end{equation}
which implies $c = 1$. Hence, we have proven $\mtheta_1 = \mtheta'_1$ provided that $z_1 = z'_1$.

@@

\section*{Proof of Theorem~\ref{thm:stats}}

@@

\section*{Proof of Theorem~\ref{thm:comp}}

@@

\section*{Proof of Theorem~\ref{thm:initial}}
\begin{proof}[Proof of Theorem 4] We prove Theorem~\ref{thm:initial} under the symmetric dTBM~\eqref{eq:model_tensor} with parameters $(z, \tS, \mtheta)$. We drop the subscript $k$ in the matricizations $\mM_k, \mX_k, \mS_k$. For simplicity, let $\hat z$ denote the output, $\hat z^{(0)}$, of Sub-algorithm~\hyperref[alg:main]{1}.

@@

\section*{Proof of Theorem~\ref{thm:refinement}}

@@

    To bound $\sin \beta_{k1}$, recall Definition~\ref{def:stable} that for any cluster assignment $\bar z$ in the $\varepsilon$-neighborhood of true $z$,
    \begin{equation}
        \mp(\bar z)=(|\bar z^{-1}(1)|, \ldots,|\bar z^{-1}(r)|)^T,\quad \mp_{\mtheta}(\bar z)=(\onormSize{}{\mtheta_{\bar z^{-1}(1)}}_1,\ldots,\onormSize{}{\mtheta_{\bar z^{-1}(r)}}_1)^T.
    \end{equation}
    Note that we have $\ell^{(t)} \leq \frac{L^{(t)}}{\Delta_{\min}^2} \lesssim \log^{-1}(p) $ by Condition~\ref{cond:origin} and Lemma~\ref{lem:mis}. With the locally linear stability assumption, we have 
    \begin{equation}
         \sin (\mp(z^{(t)}), \mp_{\mtheta}(z^{(t)})) \lesssim \frac{L^{(t)}}{\Delta_{\min}}. 
    \end{equation}
    Note that $\text{diag}(\mp(z^{(t)})) \mD = \text{diag}(\mp_{\mtheta}(z^{(t)})) \tilde \mD$, and $\sin(\ma,\mb) =\min_{c \in \bbR} \frac{ \onormSize{}{\ma - c\mb} }{\onormSize{}{\ma}}$ for vectors $\ma,\mb$ of same dimension. Let $c_0 = \argmin_{c \in \bbR} \frac{\onormSize{}{\mp(z^{(t)}) - c\mp_{\mtheta}(z^{(t)})}}{\onormSize{}{\mp(z^{(t)})}}$. Then, we have
    \begin{align}
       \min_{c \in \bbR} \onormSize{}{ \mD - c \tilde \mD}_F &\leq  \onormSize{}{ \mI_r - c_0 \text{diag}(\mp(z^{(t)})) \text{diag}^{-1}(\mp_{\mtheta}(z^{(t)}))}_F \onormSize{}{\mD}_F\\
       &\lesssim \frac{ \onormSize{}{ \mp(z^{(t)}) - c_0\mp_{\mtheta}(z^{(t)})} }{\min_{a \in [r]} \onormSize{}{\mtheta_{z^{(t),-1}(a)} }_1  } \\
       & = \frac{\onormSize{}{ \mp(z^{(t)})} }{\min_{a \in [r]} \onormSize{}{\mtheta_{z^{(t),-1}(a)} }_1 } \sin (\mp(z^{(t)}), \mp_{\mtheta}(z^{(t)}))\\
       & \lesssim \frac{L^{(t)}}{\Delta_{\min}},
    \end{align}
    where the last inequality follows from Lemma~\ref{lem:membership}, the constraint $\min_{i \in [p]}\theta(i) \geq c>0$, $\onormSize{}{ \mp(z^{(t)})} \lesssim p$ and $\min_{a \in [r]} \onormSize{}{\mtheta_{z^{(t),-1}(a)} }_1  \gtrsim p$. By the geometry property of trigonometric function, we have
    \begin{align}\label{eq:b1}
        \sin \beta_{k1} &= \min_{c \in \bbR} \frac{\onormSize{}{\mD_{:b}^{T} \mS \mI_r^{\otimes k-1} \otimes (\mD - c \tilde \mD) \otimes \mD^{\otimes K-k - 1} } }{ \onormSize{}{\mD_{:b}^{T} \mS \mI_r^{\otimes k-1} \otimes \mD^{\otimes K-k}} }\\
        &\leq \frac{ \onormSize{}{\mD_{:b}^{T} \mS} \onormSize{}{\mD - c_0 \tilde \mD}_{\sigma} \onormSize{}{\mD}_{\sigma}^{K-k-1} }{\onormSize{}{\mD_{:b}^{T} \mS} \lambda_r^{K-k}(\mD) }\\
        & \lesssim  \onormSize{}{\mD - c_0 \tilde \mD}_{F}\\
        & \lesssim \frac{L^{(t)}}{\Delta_{\min}},
    \end{align}
    where the second inequality follows from the singular property of $\mD$ in \eqref{eq:singular_Du}, \eqref{eq:singular_Dl} and the constraint of $\mS$ in \eqref{eq:family}.

>>>>

\documentclass[11pt]{article}
\usepackage{graphicx}
\usepackage{subfigure}
\usepackage{caption}
\usepackage[hidelinks]{hyperref}
\usepackage[margin=1in]{geometry}
% \geometry{
%  left=30mm,
%  right=30mm
%  }

@@

\usepackage{amsmath,amssymb,amsthm,bm,mathtools}
\usepackage{algorithm}
\usepackage{dsfont,multirow,hyperref,setspace,enumerate}
\hypersetup{colorlinks,linkcolor=blue,citecolor=blue,urlcolor={black}}
\usepackage{lscape}
\usepackage{afterpage}
\usepackage{hyperref}

@@

\usepackage{textcomp}
\usepackage{stfloats}
\usepackage{verbatim}
\usepackage{array}

@@

\usepackage{appendix}
\usepackage{fullpage} 
\usepackage{microtype}
\usepackage{wrapfig}
\mathtoolsset{showonlyrefs}

@@

\newcommand{\distgap}{D_{\ell_2}}
\newcommand{\Mat}{\text{Mat}}

@@

\newcommand\Algphase[1]{%
\vspace*{-.7\baselineskip}\Statex\hspace*{\dimexpr-\algorithmicindent-2pt\relax}\rule{\textwidth}{0.4pt}%
\Statex\hspace*{-\algorithmicindent}\textbf{#1}%
\vspace*{-.7\baselineskip}\Statex\hspace*{\dimexpr-\algorithmicindent-2pt\relax}\rule{\textwidth}{0.4pt}%
}
\usepackage{caption}
\def\fixme#1#2{\textbf{\color{red}[FIXME (#1): #2]}}

\title{Multiway Spherical Clustering via Degree-Corrected\\ Tensor Block Models\footnote{Part of this paper was presented at 25th International Conference on Artificial Intelligence and Statistics (AISTATS).}}
\date{}
\author{%
%\small
Jiaxin Hu \\
University of Wisconsin -- Madison\\
\texttt{jhu267@wisc.edu} \\
\and
Miaoyan Wang \\
University of Wisconsin -- Madison\\
\texttt{miaoyan.wang@wisc.edu}\\
}

@@

\maketitle
%\doublespacing

@@

\textbf{Keywords:} tensor clustering, degree correction, statistical-computational efficiency, human brain connectome networks

\section{Introduction}\label{sec:intro}
Multiway arrays have been widely collected in various fields including social networks \citep{anandkumar2014tensor}, neuroscience \citep{wang2017bayesian}, and computer science \citep{koniusz2016sparse}. Tensors effectively represent the multiway data and serve as the foundation in higher-order data analysis. One data example is from multi-tissue multi-individual gene expression study~\citep{wang2019three,hore2016tensor}, where the data tensor consists of expression measurements indexed by (gene, individual, tissue) triplets. Another example is \emph{hypergraph} network~\citep{ghoshdastidar2017uniform,ghoshdastidar2017consistency,ahn2019community,ke2019community} in social science. A $K$-uniform hypergraph can be naturally represented as an order-$K$ tensor, where each entry indicates the presence of $K$-way hyperedge among nodes (a.k.a.\ entities). In both examples, identifying the similarity among tensor entities is important for scientific discovery.

@@

Classical tensor block model suffers from drawbacks to model real world data in spite of the popularity. The key underlying assumption of block model is that all nodes in the same community are exchangeable; i.e., the nodes have no individual-specific parameters apart from the community-specific parameters. However, the exchangeability assumption is often non-realistic. Each node may contribute to the data variation by its own multiplicative effect. We call the unequal node-specific effects the \emph{degree heterogeneity}. Such degree heterogeneity appears commonly in social networks. Ignoring the degree heterogeneity may seriously mislead the clustering results. For example, regular block model fails to model the member affiliation in the Karate Club network~\citep{bickel2009nonparametric} without addressing degree heterogeneity.

@@

\begin{itemize}[wide]
    \item \textit{Block model for clustering.} Block models such as stochastic block model (SBM) and degree-corrected SBM have been widely used for matrix clustering problems. The theoretical properties and algorithm performance for matrix block models have been well-studied~\citep{gao2018community}; see the review paper~\citep{abbe2017community} and the references therein. However, The tensor counterparts are relatively less understood.

    \item  \textit{Tensor block model.} The (non-degree) tensor block model (TBM) is a higher-order extension of SBM, and its statistical-computational properties are investigated in recent literatures~\citep{wang2019multiway, han2020exact, ghoshdastidar2017consistency}. Extending results from non-degree to degree-corrected model is highly challenging. Our dTBM parameter space is equipped with angle-based similarity and nuisance degree parameters. The extra complexity makes the Cartesian coordinates based analysis~\citep{han2020exact} non-applicable to our setting. Towards this goal, we have developed a new polar coordinates based analysis to control the model complexity. We also develop a new angle-based iteration algorithm to achieve optimal clustering rates \emph{without the need of estimating nuisance degree parameters}.

    \item \textit{Degree-corrected block model.} The hypergraph degree-corrected block model (hDCBM) and its variant have been proposed in the literature~\citep{ke2019community, yuan2018testing}. For this popular model, however, the optimal statistical-computational rates remain an open problem. Our main contribution is to provide a sharp statistical and computational critical phase transition in dTBM literature. In addition, our algorithm results in a faster \emph{exponential} error rate, in contrast to the \emph{polynomial} rate in~\cite{ke2019community}. The original hDCBM~\citep{ke2019community} is designed for binary observations only, and we extend the model to both continuous and binary observations. We believe our results are novel and helpful to the community. See Fig~\ref{fig:phase_axis} for overview of our results.

@@

Assume there exist $r$ disjoint communities among $p$ nodes, and the connection probabilities depend on the community assignments and node-specific parameters. 
Then, the equation~\eqref{eq:model_tensor} models $\bbE \tY$ with unknown degree heterogeneity $\mtheta$ and sub-Gaussianity parameter $\sigma^2 = 1/4$.

@@

\Algphase{Sub-algorithm 2: Angle-based iteration}
\INPUT Observation $\tY \in \bbR^{p \times \cdots \times p}$, initialization $z^{(0)} \colon [p]\to[r]$ from Sub-algorithm 1, iteration number $T$.
\For {$t = 0$ to $T-1$}
\State Update the block tensor $\tS^{(t)}$ via
$\tS^{(t)} (a_1,...,a_K)= \text{Ave} \{\tY(i_1,\ldots,i_K): z^{(t)}(i_k) = a_k, k \in [K]\}.$
\State Calculate reduced tensor $\tY^{\text{d}} \in \bbR^{p \times r \times \cdots \times r}$ via
\begin{align}
    &\tY^{\text{d}}(i,a_2,\ldots,a_K) 
    = 
\text{Ave}\{\tY(i,i_2,\ldots,i_K): z^{(t)}(i_k) = a_k, k \neq 1 \}.
\end{align}

@@

\begin{defn}[Locally linear stability] \label{def:stable}
Define the $\varepsilon$-neighborhood of $z$ by $\tN(z,\varepsilon)=\{\bar z\colon \ell(\bar z, z) \leq \varepsilon\}$. Let $\bar z\colon[p]\to [r]$ be a clustering function. We define two vectors associated with $\bar z$,
\begin{align}
    &\mp(\bar z)=(|\bar z^{-1}(1)|, \ldots,|\bar z^{-1}(r)|)^T, \quad \mp_{\mtheta}(\bar z)=(\onormSize{}{\mtheta_{\bar z^{-1}(1)}}_1,\ldots,\onormSize{}{\mtheta_{\bar z^{-1}(r)}}_1)^T.
\end{align}

@@

\begin{thm}[Error for angle-based iteration]\label{thm:refinement} Consider the setup as in Theorem~\ref{thm:initial}. Suppose 
$r=\tO(1)$ and $\text{SNR} \geq \tilde C p^{-K/2}\log p$ for some sufficiently large constant $\tilde C$. Assume the local linear stability of degree holds in the neighborhood $\tN(z,\varepsilon)$ for all $\varepsilon \leq E_0$ and some $E_0 \geq \check{C}\log^{-1}p $ with some positive constant $\check{C}$.  Let $z^{(t)}$ denote the $t$-th iteration output in Sub-algorithm~\hyperref[alg:main]{2} with initialization $z^{(0)}$ from Sub-algorithm~\hyperref[alg:main]{1}. With probability going to 1, there exists a contraction parameter $\rho \in (0,1)$ such that 
\begin{align}\label{eq:final}
    \ell(z, \hat z^{(t+1)}) \lesssim &\ \KeepStyleUnderBrace{
   \text{SNR}^{-1}
    \exp\of{- \frac{p^{K-1}\text{SNR}}{r^{K-1}}}}_{\substack{\text{statistical error}} }+ \KeepStyleUnderBrace{ \rho^t \ell(z, z^{(0)}). }_{\substack{\text{computational error}}}
\end{align}
\end{thm}
From the conclusion~\eqref{eq:final}, we find that the iteration error is decomposed into two parts: statistical error and computational error. The statistical error is unavoidable with noisy data regardless $t$, whereas the computational error decays in an exponential rate as the number of iterations $t \rightarrow \infty$.

@@

{\bf Extension for general dTBMs.} Our two-stage algorithm is able to be extended for the general (asymmetric) dTBMs. Specifically, in the Sub-algorithm~\hyperref[alg:main]{1}, we make the following changes: (1)
Replace the matrcization $\mat(\tY)$ by $\mat_k(\tY)$; (2) Repeat the Steps 1-5 with mode-specified cluster number $r_k$; (3) Obtain the collection initialization $ \{z^{(0)}_k\}_{k=1}^K$. In the Sub-algorithm~\hyperref[alg:main]{2}, we make the following changes: (1) Take the collection $\{z^{(0)}_k\}_{k=1}^K$ as input, and update the block tensor $\tS^{(t)}$ with the collection $\{ z_k^{(t)}\}_{k=1}^K$, $
        \tS^{(t)}(a_1,\ldots, a_K) =  \text{Ave} \{\tY(i_1,\ldots,i_K): z^{(t)}_k(i_k) = a_k, k \in [K]\}$;
(2) Calculate reduced tensor $\tY^{\text{d}}_k$ for each mode via
\begin{align}
    &\tY^{\text{d}}_k(a_1,\ldots,a_{k-1}, i ,a_{k+1},\ldots,a_K) 
    = \text{Ave}\{\tY(i_1,\ldots,i_{k-1},i,i_{k+1},\ldots,i_K): z^{(t)}(i_j) = a_j, j \neq k \}; 
\end{align}

@@

\begin{table}[hbt]
\centering
    \begin{tabularx}{\columnwidth}{c *{8}{Y}}
    \toprule
    Settings & \multicolumn{2}{c}{\small  $p = 50, \sigma^2 = 0.25$} & \multicolumn{2}{c}{\small $p = 50, \sigma^2 = 1$} & \multicolumn{2}{c}{\small  $p = 80, \sigma^2 = 0.25$} & \multicolumn{2}{c}{\small  $p = 80, \sigma^2 = 1$}\\
    \cmidrule(lr){2-3} \cmidrule(l){4-5} \cmidrule(l){6-7} \cmidrule(l){8-9}
         True cluster number $r$ &  \small 2 & \small 4 &  \small2  &  \small 4 &  \small2  & \small 4 &  \small 2 & \small 4  \\
         \midrule
         Estimated cluster number $\hat r$ &   \small 2(0)  &  \small 3.9(0.2)&  \small 2(0)    &   \small 3.1(0.5) &  \small 2(0)    &  \small 4(0)   &  \small 2(0)    &  \small 3.9(0.3)   \\
     \bottomrule
    \end{tabularx}
    \caption{Estimated cluster number given by BIC criterion under the low noise level $(\sigma^2 = 0.25)$ and high noise level $(\sigma^2 = 0.5)$ settings. Numbers in parentheses are standard deviations of $\hat r$ over 30 replications.}
    \label{tab:select}
\end{table}

@@

\newpage
\appendix
\section*{Appendices}

@@

\section{Notation}
Before the proofs, we first introduce the notation used throughout the appendix and the generalized dTBM without symmetric assumptions. The parameter space and minimal gap assumption are also extended for the generalized dTBM.

@@

Consider an arbitrary $j \in [p_1]$ such that $z_1(j) = a$. Then for all the nodes $i \in  z_1^{-1}(a)$ in the same cluster of $j$, we have 
\begin{equation}
    \frac{\mX_{1,z_1(i):}}{\mX_{1,z_1(j):}} = \frac{\mX'_{1,z_1(i):}}{\mX'_{1,z_1(j):}}, \text{ which implies }  \frac{\theta_1(j)}{\theta_1(i)} = \frac{\theta'_1(j)}{\theta'_1(i)}.\label{eq:theta_uniq}
\end{equation}
Let $\theta'_1(j) = c\theta_1(j)$ for some positive constant $c$. By equation~\eqref{eq:theta_uniq}, we have $\theta'_1(i) = c \theta_1(i)$ for all $ i \in  z_1^{-1}(a)$. By the constraint $(\{z_k\}, \tS', \{\mtheta'_k\}) \in \tP(\{r_k\})$, we have 
\begin{equation}
    \sum_{j \in z_1^{-1}(a)} \theta'_1(j) = c \sum_{j \in z_1^{-1}(a)} \theta_1(j) = 1,
\end{equation}
which implies $c = 1$. Hence, we have proven $\mtheta_1 = \mtheta'_1$ provided that $z_1 = z'_1$.

@@

\section{Proof of Theorem~\ref{thm:stats}}

@@

\section{Proof of Theorem~\ref{thm:comp}}

@@

\section{Proof of Theorem~\ref{thm:initial}}
\begin{proof}[Proof of Theorem 4] We prove Theorem~\ref{thm:initial} under the symmetric dTBM~\eqref{eq:model_tensor} with parameters $(z, \tS, \mtheta)$. We drop the subscript $k$ in the matricizations $\mM_k, \mX_k, \mS_k$. For simplicity, let $\hat z$ denote the output, $\hat z^{(0)}$, of Sub-algorithm~\hyperref[alg:main]{1}.

@@

\section{Proof of Theorem~\ref{thm:refinement}}

@@

    To bound $\sin \beta_{k1}$, recall Definition~\ref{def:stable} that for any cluster assignment $\bar z$ in the $\varepsilon$-neighborhood of true $z$,
    \begin{equation}
        \mp(\bar z)=(|\bar z^{-1}(1)|, \ldots,|\bar z^{-1}(r)|)^T,\quad \mp_{\mtheta}(\bar z)=(\onormSize{}{\mtheta_{\bar z^{-1}(1)}}_1,\ldots,\onormSize{}{\mtheta_{\bar z^{-1}(r)}}_1)^T.
    \end{equation}
    Note that we have $\ell^{(t)} \leq \frac{L^{(t)}}{\Delta_{\min}^2} \leq \frac{\bar C}{\tilde C} r \log^{-1}(p) $ by Condition~\ref{cond:origin} and Lemma~\ref{lem:mis}. Then,  with the locally linear stability assumption, the $\mtheta$ is $\ell^{(t)}$-locally linearly stable; i.e.,
    \begin{equation}
         \sin (\mp(z^{(t)}), \mp_{\mtheta}(z^{(t)})) \lesssim \frac{L^{(t)}}{\Delta_{\min}}. 
    \end{equation}
    Note that $\text{diag}(\mp(z^{(t)})) \mD = \text{diag}(\mp_{\mtheta}(z^{(t)})) \tilde \mD$, and $\sin(\ma,\mb) =\min_{c \in \bbR} \frac{ \onormSize{}{\ma - c\mb} }{\onormSize{}{\ma}}$ for vectors $\ma,\mb$ of same dimension. Let $c_0 = \argmin_{c \in \bbR} \frac{\onormSize{}{\mp(z^{(t)}) - c\mp_{\mtheta}(z^{(t)})}}{\onormSize{}{\mp(z^{(t)})}}$. Then, we have
    \begin{align}
       \min_{c \in \bbR} \onormSize{}{ \mD - c \tilde \mD}_F &\leq  \onormSize{}{ \mI_r - c_0 \text{diag}(\mp(z^{(t)})) \text{diag}^{-1}(\mp_{\mtheta}(z^{(t)}))}_F \onormSize{}{\mD}_F\\
       &\lesssim \frac{ \onormSize{}{ \mp(z^{(t)}) - c_0\mp_{\mtheta}(z^{(t)})} }{\min_{a \in [r]} \onormSize{}{\mtheta_{z^{(t),-1}(a)} }_1  } \\
       & = \frac{\onormSize{}{ \mp(z^{(t)})} }{\min_{a \in [r]} \onormSize{}{\mtheta_{z^{(t),-1}(a)} }_1 } \sin (\mp(z^{(t)}), \mp_{\mtheta}(z^{(t)}))\\
       & \lesssim \frac{L^{(t)}}{\Delta_{\min}},
    \end{align}
    where the last inequality follows from Lemma~\ref{lem:membership}, the constraint $\min_{i \in [p]}\theta(i) \geq c>0$, $\onormSize{}{ \mp(z^{(t)})} \lesssim p$ and $\min_{a \in [r]} \onormSize{}{\mtheta_{z^{(t),-1}(a)} }_1  \gtrsim p$. By the geometry property of trigonometric function, we have
    \begin{align}\label{eq:b1}
        \sin \beta_{k1} &= \min_{c \in \bbR} \frac{\onormSize{}{\mD_{:b}^{T} \mS \mI_r^{\otimes k-1} \otimes (\mD - c \tilde \mD) \otimes \mD^{\otimes K-k - 1} } }{ \onormSize{}{\mD_{:b}^{T} \mS \mI_r^{\otimes k-1} \otimes \mD^{\otimes K-k}} }\\
        &\leq \frac{ \onormSize{}{\mD_{:b}^{T} \mS} \onormSize{}{\mD - c_0 \tilde \mD}_{\sigma} \onormSize{}{\mD}_{\sigma}^{K-k-1} }{\onormSize{}{\mD_{:b}^{T} \mS} \lambda_r^{K-k}(\mD) }\\
        & \lesssim  \onormSize{}{\mD - c_0 \tilde \mD}_{F}\\
        & \lesssim \frac{L^{(t)}}{\Delta_{\min}},
    \end{align}
    where the second inequality follows from the singular property of $\mD$ in \eqref{eq:singular_Du}, \eqref{eq:singular_Dl} and the constraint of $\mS$ in \eqref{eq:family}.
